Dust trapping in thin-ringed protoplanetary disks

### 2018-08-13 projects/dullemond/paper_dusttraps/ms.tex
Slight update of dust trapping paper.

\begin{figure*}
\centerline{\includegraphics[width=0.85\textwidth]{gallery_profiles.pdf}}
\caption{\label{fig-obs-profiles}The linear brightness temperature in
  band 6 of the five disks in our sample which have the most pronounced rings.
  The vertical axis is logarithmic to better show the contrast.
  The eight highest contrast rings are fitted by a Gaussian profile, shown
  as the solid inverse parabolas. The dotted inverse parabolas are Gaussians
  with the width of the ALMA beam. For a detailed description of these
  data, \paperdsharpguzmann{} for AS 209, \paperdsharpisella{}
  for HD 163296, and \paperdsharphuangrings{} for the rest.}
\end{figure*}

@@

{\bf [ADD APPENDIX WITH DETAILS OF FITS AND ERROR ESTIMATES]}

\begin{figure*}
\centerline{\includegraphics[width=0.85\textwidth]{gaussfits_allrings.pdf}}
\caption{\label{fig-obs-gaussfits}Gaussian fits to the eight rings of this
  paper. The blue curves are the observations, the orange curves are the best fit
  Gaussian profiles. The ``fit range'' bar shows the radial range within which
  the Gauss was fitted to the data. The fit range was chosen to fit the part of
  the curve that, by eye, most resembles a Gaussian. The ``beam'' bar shows the
  FWHM beam size of the observations. The grey band around the blue curve shows
  the estimated uncertainty of the data.}
\end{figure*}

\begin{figure*}
  \centerline{
    %\includegraphics[width=0.3\textwidth]{AS_209_doublegauss.pdf}
    \includegraphics[width=0.42\textwidth]{HD_163296_doublegauss.pdf}
    \includegraphics[width=0.42\textwidth]{HD_143006_doublegauss.pdf}
  }
  \caption{\label{fig-obs-doublegauss}The sum of the two Gaussian fits
    for the two sources with two partly overlapping rings: HD 163296 and
    HD 143006.}
\end{figure*}

@@

The observed rings are the result of the thermal emission of a dust ring convolved
with the ALMA beam. To obtain the width of the
underlying dust ring we have to deconvolve. Assuming a Gaussian beam and a
Gaussian dust ring, we can use the rule of the convolution of two Gaussians, and
obtain the width $w_d$ of the dust ring
\begin{equation}\label{eq-simple-deconvolve-gauss}
w_d=\sqrt{\sigma^2-b^2}
\end{equation}
where $b$ is the beam width expressed as standard deviation in units of
$\mathrm{au}$ ($b=d_{\mathrm{pc}}b_{\mathrm{fwhm,as}}/2.355$, where
$b_{\mathrm{fwhm,as}}$ is the beam full-width at half-maximum (FWHM) in units of
arcseconds, and $d_{\mathrm{pc}}$ is the distance to the source in units of
parsec). In using Eq.~(\ref{eq-simple-deconvolve-gauss}) we have ignored the
elliptical shape of the beam as well as the inclination of the disk.

@@

The width of the dust rings
in units of the local pressure scale height $h_p$ (see Table
\ref{tab-gauss-params}) can only be roughly estimated, because we do not know
the disk midplane temperature. Our estimate of the midplane disk temperature
is made using the following simple irradiated flaring disk recipe:
\begin{equation}\label{eq-disk-temperature-model}
T_{\mathrm{d}}(r) = \left(\frac{\tfrac{1}{2}\varphi L_{*}}{4\pi r^2\sigma_{\mathrm{SB}}}\right)^{1/4}
\end{equation}
where $\sigma_{\mathrm{SB}}$ is the Stefan-Boltzmann constant and $\varphi$ is
the so-called flaring angle \citep[e.g.][]{1997ApJ...490..368C,
  1998ApJ...500..411D, 2001ApJ...560..957D}. The factor of $1/2$ in front of the
flaring angle $\varphi$ originates from the consideration that the stellar
radiation heats the surface layer of the disk, which then radiates half of that
energy away and half into the deeper regions of the disk. Only the latter half
goes into the energy balance equation between irradiation and radiative cooling
of the disk interior. We take the flaring angle to be $\varphi=0.02$ which is an
estimate based on typical values from models. In using this simple recipe we
assume that the optical and infrared dust opacity is dominated by a population
of small grains that is vertically more extended than the larger dust grains
seen in the ALMA images. The two dust populations may
be largely decoupled (see e.g.\ the BL-series of models of
\citet{2004A&A...417..159D} as an illustration). The value of $\varphi=0.02$ is
nothing more than an educated guess, so its value can easily be wrong by a
factor of four or so. But the fact that the pressure scale height goes as
$T_{\mathrm{d}}^{1/2}\propto \varphi^{1/8}$ means that such an error only
affects $h_p$ mildly.

@@

For these observations the FWHM beam size for each ring is listed in Table
\ref{tab-gauss-params}. These are azimuthal average beam size values according
to Eq.~(\ref{eq-def-average-beam}), based on the values listed in
\paperdsharphuangrings{}. This number can be compared to the FWHM ring width in
Table \ref{tab-gauss-params}, and it shows that the widths of all the rings are
spatially resolved. Some are only marginally resolved (AS 209 rings 1 and 2, GW
Lup and the ring 1 of HD 143006), but the data are inconsistent with radially
unresolved rings. For AS 209 This is confirmed by more detailed analysis in the
uv-plane by \paperdsharpguzmann{}. {\bf CHECK IF HUANG DOES THIS, TOO.}

@@

By replacing $I_\nu^{\mathrm{lb,dec}}(r)$ with the Gaussian fit
$I_\nu^{\mathrm{lb,gauss,dec}}(r)$ given by
Eq.~(\ref{eq-gauss-lin-br-temp-deconv}) we obtain the corresponding
$\Sigma_d^{\mathrm{gauss}}(r)$ from
Eq.~(\ref{eq-optthin-conversion-linbright-sigmad}). From this Gaussian
model we can derive the total dust mass trapped in the ring:
\begin{equation}
  M_d = \int_0^\infty 2\pi r \Sigma_d(r) dr \simeq
  \frac{(2\pi)^{3/2} r_0\, A\,\sigma}{\kappa_\nu^{\mathrm{abs}}\, B_\nu^{\mathrm{lb}}(T_d)}
\end{equation}
where we used the identity $A\,\sigma=A_{\mathrm{dec}}\,w_d$.

@@

Given the amplitude of the deconvolved Gauss fit $A_{\mathrm{dec}}$ (see
Eq.~\ref{eq-gauss-lin-br-temp-deconv}), we can derive the optical depth
$\tau_\nu^{\mathrm{peak}}$ of the ring at its peak at $r=r_0$:
\begin{equation}\label{eq-tau-estimate}
\tau_\nu^{\mathrm{peak}} = -\ln\left(1-\frac{A_{\mathrm{dec}}}{B_{\nu}^{\mathrm{lb}}(T_d)}\right)
\end{equation}
This allows us to verify if our assumption of optical thinness is justified. The
results are listed in Table \ref{tab-gauss-params}.  We conclude that the
optically thin assumption is not entirely wrong, but not quite right either. The
estimated dust surface density $\Sigma_d$ and dust mass $M_d$ derived from the
optically thin assumption ( Eqs.~\ref{eq-optthin-conversion-linbright-sigmad}
and {\bf XXX}, respectively) are only mildly affected by the optical depth
effects.

@@

\subsection{The effect of scattering albedo}
If the dust grains have a radius $a$ comparable to the wavelength of our
observations, the scattering albedo can be quite high. This means that the
absorption optical depth can be substantially lower than unity, even if the full
extinction optical depth (absorption plus scattering) is unity or larger.  The
near-unity extinction found in HD 163296 by \paperdsharpisella{} from the CO
maps could thus be compatible with the sub-unity absorption optical depth we
derived in out Gaussian fitting procedure (see Table \ref{tab-gauss-params}).

@@

\section{The rings as dust traps}
\label{sec-rings-as-dust-traps}
%
Given the well separated narrow rings in the above sources, we can analyze these
rings on a one-by-one basis. The hypothesis we are going to test is that the
rings are caused by dust trapping in axisymmetric pressure bumps. For simplicity
we will assume that the radial gas pressure profile is fixed in time, and there
is no back reaction of the dust onto the gas. The pressure bump is assumed to be
so strong that the dust trapping in these rings is nearly perfect: no dust
escapes. We then expect that the dust distribution finds an equilibrium between
dust drift and turbulent spreading. Low turbulence will lead to thinner dust
rings than high turbulence. For simplicity we shall assume the gas pressure bump
to have a Gaussian radial profile with its peak at radius $r_0$ and width $w\ll
r_0$. This problem can be solved analytically.

@@

{\bf Complete this section. Add an example of a corner plot (only one is
presumably enough) and provide a more detailed table of error estimates.}

The fitting was done in the
image plane, in order to make it possible to fit to a limited radial range,
i.e.\ selecting a single ring.

@@

Thes results show that up to $\tau_\nu^{\mathrm{peak}}\lesssim 1$ the
differences between the best-fit Gauss and the real curve are at most $\sim
1\cdots 2$ \%, depending on the beam size. The beam size has most influence
on ${\cal F}$, but less on the other parameters. That is because many of the
effects of beam smearing are cancelled out because we know the beam size and
can deconvolve. But the beam smearing does make a non-Gaussian profile more
Gaussian-shaped, hence the stronger effect on ${\cal F}$.

>>>>

\begin{figure*}
\centerline{\includegraphics[width=0.85\textwidth]{gallery_profiles.pdf}}
\caption{\label{fig-obs-profiles}The linear brightness temperature in
  band 6 of the five disks in our sample which have the most pronounced rings.
  The vertical axis is logarithmic to better show the contrast.
  The eight highest contrast rings are fitted by a Gaussian profile, shown
  as the solid inverse parabolas. The dotted inverse parabolas are Gaussians
  with the width of the ALMA beam. For a detailed description of these
  data, \paperdsharpguzmann{} for AS 209, \paperdsharpisella{}
  for HD 163296, and \paperdsharphuangrings{} for the rest.
{\bf TODO: y-labels}}
\end{figure*}

@@

\begin{figure*}
\centerline{\includegraphics[width=0.85\textwidth]{gaussfits_allrings.pdf}}
\caption{\label{fig-obs-gaussfits}Gaussian fits to the eight rings of this
  paper. The blue curves are the observations, the orange curves are the best fit
  Gaussian profiles. The ``fit range'' bar shows the radial range within which
  the Gauss was fitted to the data. The fit range was chosen to fit the part of
  the curve that, by eye, most resembles a Gaussian. The ``beam'' bar shows the
  FWHM beam size of the observations. The grey band around the blue curve shows
  the estimated uncertainty of the data. {\bf TODO: y-labels}}
\end{figure*}

\begin{figure*}
  \centerline{
    %\includegraphics[width=0.3\textwidth]{AS_209_doublegauss.pdf}
    \includegraphics[width=0.42\textwidth]{HD_163296_doublegauss.pdf}
    \includegraphics[width=0.42\textwidth]{HD_143006_doublegauss.pdf}
  }
  \caption{\label{fig-obs-doublegauss}The sum of the two Gaussian fits
    for the two sources with two partly overlapping rings: HD 163296 and
    HD 143006. {\bf TODO: y-labels}}
\end{figure*}

@@

The observed rings are the result of the thermal emission of a dust ring convolved
with the ALMA beam. To obtain the width of the
underlying dust ring we have to deconvolve. Assuming a Gaussian beam and a
Gaussian dust ring, we can use the rule of the convolution of two Gaussians, and
obtain the width $w_d$ of the dust ring
\begin{equation}\label{eq-simple-deconvolve-gauss}
w_d=\sqrt{\sigma^2-b^2}
\end{equation}
where $b$ is the beam width expressed as standard deviation in units of
$\mathrm{au}$. It is computed from $b_{\mathrm{fwhm,as}}$ (the beam full-width
at half-maximum (FWHM) in units of arcseconds) through the equation
$b=d_{\mathrm{pc}}b_{\mathrm{fwhm,as}}/2.355$, where $d_{\mathrm{pc}}$ is the
distance to the source in units of parsec. In using
Eq.~(\ref{eq-simple-deconvolve-gauss}) we have ignored the elliptical shape of
the beam as well as the inclination of the disk.

@@

The width of the dust rings in units of the local pressure scale height $h_p$
(see Table \ref{tab-gauss-params}) can only be roughly estimated, because we do
not know $h_p$ well enough due to our lack of information about the disk midplane
temperature. Our estimate of the midplane disk temperature is made using the
following simple irradiated flaring disk recipe:
\begin{equation}\label{eq-disk-temperature-model}
T_{\mathrm{d}}(r) = \left(\frac{\tfrac{1}{2}\varphi L_{*}}{4\pi r^2\sigma_{\mathrm{SB}}}\right)^{1/4}
\end{equation}
where $\sigma_{\mathrm{SB}}$ is the Stefan-Boltzmann constant and $\varphi$ is
the so-called flaring angle \citep[e.g.][]{1997ApJ...490..368C,
  1998ApJ...500..411D, 2001ApJ...560..957D}. The factor of $1/2$ in front of the
flaring angle $\varphi$ originates from the consideration that the stellar
radiation heats the surface layer of the disk, which then radiates half of that
energy away and half into the deeper regions of the disk. Only the latter half
goes into the energy balance equation between irradiation and radiative cooling
of the disk interior. We take the flaring angle to be $\varphi=0.02$ which is an
estimate based on typical values from models. In using this simple recipe we
assume that the optical and infrared dust opacity is dominated by a population
of small grains that is vertically more extended than the larger dust grains
seen in the ALMA images. The two dust populations may
be largely decoupled (see e.g.\ the BL-series of models of
\citet{2004A&A...417..159D} as an illustration). The value of $\varphi=0.02$ is
nothing more than an educated guess, so its value can easily be wrong by a
factor of four or so. But the fact that the pressure scale height goes as
$T_{\mathrm{d}}^{1/2}\propto \varphi^{1/8}$ means that such an error only
affects $h_p$ mildly.

@@

For these observations the FWHM beam size for each ring is listed in Table
\ref{tab-gauss-params}. These are azimuthal average beam size values according
to Eq.~(\ref{eq-def-average-beam}), based on the values listed in
\paperdsharphuangrings{}. This number can be compared to the FWHM ring width in
Table \ref{tab-gauss-params}, and it shows that the widths of all the rings are
spatially resolved. Some are only marginally resolved (AS 209 rings 1 and 2, GW
Lup and the ring 1 of HD 143006), but the data are inconsistent with radially
unresolved rings. For AS 209 this is confirmed by more detailed analysis in the
uv-plane by \paperdsharpguzmann{}. {\bf CHECK IF HUANG DOES THIS, TOO.}

@@

By replacing $I_\nu^{\mathrm{lb,dec}}(r)$ with the Gaussian fit
$I_\nu^{\mathrm{lb,gauss,dec}}(r)$ given by
Eq.~(\ref{eq-gauss-lin-br-temp-deconv}) we obtain the corresponding
$\Sigma_d^{\mathrm{gauss}}(r)$ from
Eq.~(\ref{eq-optthin-conversion-linbright-sigmad}). From this Gaussian
model we can derive the total dust mass trapped in the ring:
\begin{equation}\label{eq-dust-mass-estimate}
  M_d = \int_0^\infty 2\pi r \Sigma_d(r) dr \simeq
  \frac{(2\pi)^{3/2} r_0\, A\,\sigma}{\kappa_\nu^{\mathrm{abs}}\, B_\nu^{\mathrm{lb}}(T_d)}
\end{equation}
where we used the identity $A\,\sigma=A_{\mathrm{dec}}\,w_d$.

@@

Given the amplitude of the deconvolved Gauss fit $A_{\mathrm{dec}}$ (see
Eq.~\ref{eq-gauss-lin-br-temp-deconv}), we can derive the optical depth
$\tau_\nu^{\mathrm{peak}}$ of the ring at its peak at $r=r_0$:
\begin{equation}\label{eq-tau-estimate}
\tau_\nu^{\mathrm{peak}} = -\ln\left(1-\frac{A_{\mathrm{dec}}}{B_{\nu}^{\mathrm{lb}}(T_d)}\right)
\end{equation}
This allows us to verify if our assumption of optical thinness is justified. The
results are listed in Table \ref{tab-gauss-params}.  We conclude that the
optically thin assumption is not entirely wrong, but not quite right either. The
estimated dust surface density $\Sigma_d$ and dust mass $M_d$ derived from the
optically thin assumption (Eqs.~\ref{eq-optthin-conversion-linbright-sigmad} and
\ref{eq-dust-mass-estimate}, respectively) are only mildly affected by the
optical depth effects.

@@

\subsection{The effect of scattering albedo}
If the dust grains have a radius $a$ comparable to the wavelength of our
observations, the scattering albedo can be quite high. This means that the
absorption optical depth can be substantially lower than unity, even if the full
extinction optical depth (absorption plus scattering) is unity or larger.  The
near-unity extinction found in HD 163296 by \paperdsharpisella{} from the CO
maps could thus be compatible with the sub-unity absorption optical depth we
derived in our Gaussian fitting procedure (see Table \ref{tab-gauss-params}).

@@

\section{The rings as dust traps}
\label{sec-rings-as-dust-traps}
%
The hypothesis we are now going to test is that the
rings are caused by dust trapping in axisymmetric pressure bumps. For simplicity
we will assume that the radial gas pressure profile is fixed in time, and there
is no back reaction of the dust onto the gas. The pressure bump is assumed to be
so strong that the dust trapping in these rings is nearly perfect: no dust
escapes. We then expect that the dust distribution finds an equilibrium between
dust drift and turbulent spreading. Low turbulence will lead to thinner dust
rings than high turbulence. For simplicity we shall assume the gas pressure bump
to have a Gaussian radial profile with its peak at radius $r_0$ and width $w\ll
r_0$. This problem can be solved analytically.

@@

The radial intensity profiles (expressed as linear brightness temperature) were
extracted from the images using a procedure similar to that described by
\paperdsharphuangrings{}. This procedure involves the fitting of an ellipse to
describe the inclined ring shape, the deprojection into a ring, and the
averaging of the intensity along the ring. This averaging procedure enhances the
signal-to-noise ratio considerably, by a factor $\sqrt{N}$, where $N$ is the
number of beams that fit along the ring. We estimate the intrinsic noise simply
by computing the standard deviation along the ring. The resulting averaged
radial intensity profile $I_\nu^{\mathrm{lb}}(r)$ thus obtains also an error
estimate $\varepsilon(r)$, which is typically of the order of $\sim$1\% of
the peak intensity.

The rings display themselves as bumps in $I_\nu^{\mathrm{lb}}(r)$. We choose by eye
a radial domain around the bump where we believe a Gaussian description is justified.
The inner and outer radii of this domain are listed in Table \ref{tab-gauss-params}.
By choosing this domain we can select a specific ring to fit, which is not possible
when doing the fitting procedure in the uv-plane.

We now fit a Gaussian profile to this bump
\begin{equation}
I_\nu^{\mathrm{lb,gauss}}(r) = A\exp\left(-\frac{(r-r_0)^2}{2\sigma^2}\right)
\end{equation}
We use the code {\small\sf emcee} \citep{2013PASP..125..306F} to perform a
Markov Chain Monte Carlo procedure to find the set of parameters
$(A,r_0,\sigma)$ which have the highest likelihood. Given that the radial
sampling of $I_\nu^{\mathrm{lb,gauss}}(r)$ is about $N\simeq 70$ points per beam, we
have to multiply the error estimate of the datapoints by $\sqrt{N}$ before
feeding it into {\small\sf emcee}, otherwise the MCMC procedure would treat each
sampling point as an independent measuring point, which is not the case.

We show the corner plot of the fitting of ring 1 of AS 209 in
Fig.~\ref{fig-corner-gaussfit-as209-1}. The plots for the other rings are
similar, and are provided in the online material. {\bf TODO: PUT OTHER PLOTS
IN ONLINE MATERIAL.}

\begin{figure}
\centerline{\includegraphics[width=0.45\textwidth]{AS_209_ring1_cornerplot_gaussfit.pdf}}
\caption{\label{fig-corner-gaussfit-as209-1}The correletion plot for the Gauss
  fitting of ring 1 of AS 209, from the MCMC procedure described in Appendix
  \ref{sec-gauss-fitting-procedure}.}
\end{figure}

The most likely parameter values and their error estimates are given in
Table \ref{tab-gauss-fit-errors}.

\begin{table}
\begin{center}
\begin{tabular}{|cc|ccc|}
\hline
\hline
Source & Ring & $A$ & $r_0$ & $\sigma$ \\
\hline
AS 209     & 1 & $  3.637_{-0.047}^{+0.048}$ & $ 74.129_{-0.066}^{+0.063}$ & $  3.885_{-0.090}^{+0.098}$  \\
AS 209     & 2 & $  2.949_{-0.031}^{+0.030}$ & $120.309_{-0.068}^{+0.065}$ & $  4.651_{-0.111}^{+0.115}$  \\
Elias 24   & 1 & $  5.214_{-0.068}^{+0.066}$ & $ 76.851_{-0.092}^{+0.088}$ & $  5.104_{-0.172}^{+0.186}$  \\
HD 163296  & 1 & $  8.475_{-0.056}^{+0.056}$ & $ 67.403_{-0.055}^{+0.057}$ & $  7.370_{-0.098}^{+0.098}$  \\
HD 163296  & 2 & $  4.918_{-0.026}^{+0.027}$ & $ 99.862_{-0.061}^{+0.059}$ & $  5.741_{-0.116}^{+0.119}$  \\
GW Lup     & 1 & $  1.467_{-0.052}^{+0.055}$ & $ 85.178_{-0.259}^{+0.316}$ & $  4.882_{-0.402}^{+0.530}$  \\
HD 143006  & 1 & $  3.202_{-0.084}^{+0.082}$ & $ 41.080_{-0.169}^{+0.190}$ & $  5.133_{-0.320}^{+0.373}$  \\
HD 143006  & 2 & $  2.473_{-0.043}^{+0.044}$ & $ 64.831_{-0.248}^{+0.230}$ & $  8.029_{-0.501}^{+0.593}$  \\
\hline
\hline
\end{tabular}
\end{center}
\caption{\label{tab-gauss-fit-errors}The Gaussian fit values with their error
  estimates from the MCMC procedure described in Appendix \ref{sec-gauss-fitting-procedure}.}
\end{table}

@@

These results show that up to $\tau_\nu^{\mathrm{peak}}\lesssim 1$ the
differences between the best-fit Gauss and the real curve are at most $\sim
1\cdots 2$ \%, depending on the beam size. The beam size has most influence
on ${\cal F}$, but less on the other parameters. That is because many of the
effects of beam smearing are cancelled out because we know the beam size and
can deconvolve. But the beam smearing does make a non-Gaussian profile more
Gaussian-shaped, hence the stronger effect on ${\cal F}$.
